Using SMT engine to generate Symbolic Automata\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2017-04-21 2017-Xudong-SMT/submit-ICTAC2017.tex
finished ??

%\usepackage{amsfonts}
%\usepackage{fancyhdr}
%\usepackage{cite}
%\usepackage{ifthen}
%\usepackage{amssymb}
%\usepackage{fancyhdr}
%\usepackage{pifont}
\usepackage{stmaryrd}
\usepackage{mathtools,mathpartir}
\usepackage{proof}
%\usepackage{setspace}
%\usepackage{indentfirst}
\usepackage{amsmath,amssymb,amscd,mathrsfs}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
\usepackage{epsfig,color,subfigure,enumitem}
\newcommand{\TODO}[1]{\textcolor{red}{\textbf{[TODO:#1]}}}
\newcommand{\NOTE}[1]{\textcolor{blue}{\textbf{[NOTE:#1]}}}
\newcommand{\ERIC}[1]{\textcolor{blue}{#1}}
\newcommand{\OTvar}{\texttt}
\newcommand{\OTland}{\ \land\ }
\definecolor{airforceblue}{rgb}{0.26, 0.44, 0.56}
\newcommand{\QIN}[1]{\textcolor{airforceblue}{#1}}
\newcommand{\coloncolon}{{:\hspace{-.2ex}:}}
\makeatletter
\newcommand{\raisemath}[1]{\mathpalette{\raisem@th{#1}}}
\newcommand{\raisem@th}[3]{\raisebox{#1}{$#2#3$}}
\makeatother

@@

\title{Using SMT engine to generate Symbolic Automata\thanks{This work was partially 
funded by the Associated Team FM4CPS
  between INRIA and ECNU, Shanghai}}
\author{ Xudong Qin\inst{3}  \ \ \  Eric Madelaine\inst{1,2}} 
\institute{Univ. of Nice Sophia Antipolis, CNRS, UMR 7271, 06900 Sophia Antipolis, France
	\and INRIA Sophia Antipolis M\'edit\'erann\'ee, BP 93, 06902 Sophia Antipolis, France
\and Shanghai Key Laboratory of Trustworthy Computing, ECNU, China}
\date{}                                           % Activate to display a given date or no date

@@

\begin{abstract}
  \TODO{not good for a conference, (to be rewriten by Eric)}
  We implement the symbolic semantics of open pNets using their
  so-called ``Open Automaton'' behavioral semantics. This involve
  building predicates expressing the synchronization conditions
  allowing some combination of events in the pNet system. These
  predicates are typically built using first order logic, plus some
  predicates specific of particular action algebras. To reduce the
  complexity of the generated open automata, we use the Z3 SMT engine
  to check satisfiability of the predicates, and prune the state
  space..

@@

Parameterized Networks of Synchronized Automata (pNets) were proposed
to give a behavioral specification formalism for distributed
systems, synchronous, asynchronous, or heterogeneous. It is used in
VerCors, a platform for designing and 
verifying distributed systems, as the intermediate language for various
high-level languages. The high-level languages in VerCors formalize
each component of the 
distributed system and gives out the composition of these
components.
pNets provides the core low-level semantic formalism for VerCors, and
is made of a hierarchical composition of (value-passing) automatas,
called parameterized labelled transition systems (pLTS), where each
hierarchical level defines the possible synchronization of the lower levels.
Traditionally, pNet have been used to formalize full
defined systems or softwares. But we want also to define and reason
about incompletely defined systems, like program skeletons, operators,
or open expressions of process calculi.
The open pNet is proposed to solve the
problem of the undefined components contained in the systems using
"holes" as the process parameter dealing with these components. The
hole acts as the placeholder for the uncertain component with a set of
possible behaviors the component might conduct, presenting the "open"
property.

The pNet model was developed in a series of
works~\cite{HMZ:PDP15,henrio:Forte2016} and some examples have been
introduced showing its ability in encoding the operators from some
other algebras or  program skeletons. It is just the first step of
our long-term goals. The operational semantic of the open pNet has
already been defined, but we only had a sketch about how to compute it and
had no implementation. This operational semantic is defined as an
Open Automaton in which Open Transitions contain logical predicates
expressing the relations between the behavior of the holes, and the
global behavior of the system.

Computing the open transitions from the composition of pNet is the
main part of the open 
automaton computing and needs to satisfy the constraints expressed as
the rules of the operational semantics. So in
the computing algorithm, computing open transitions is necessary and
should have some method to keep them satisfying the constraints.
Computing open automata by hand is complicated and error prone. And
the algorithm published in \cite{henrio:Forte2016} was not precise
enough for an implementation. Our first goal was then to refine the
algorithm, and implement it in a tool.
In the long goal, we want to be able to check the equivalence between
pNets. The equivalence between pNets is "FH-bisimulation" taking the
predicate of the open transitions into account each time matching such
open transitions. In our long-term goals, the final target is to check
the property of open systems.

@@

\paragraph{Contribution}
In the article we show how:
\begin{itemize}
\item \ERIC{We have refined the open automaton generation algorithm, to make
  it implementable, and we implemented a full working prototype, within the
  VerCors platform. In the process, we had to improve significantly the
  semantics rules, and add features in the algorithm to deal the full
  model, including management of variables and assignments.}
%  \item We extended the meta-model of action expressions, and the
%    associated API, in the VerCors platform, to be able to manage
%    parameterisation over the action algebra operators.
  \item We have implemented the interaction between our algorithm and the Z3
    SMT solver, for checking satisfiability of the logical predicates
    generated by the algorithm.
    \item \ERIC{We validated the whole approach on a small use-case covering all
      important features of the model.}
\end{itemize}

@@

\TODO{Should be improved}

@@

\QIN{
For other applications, e.g. for the analyses of 
programming languages, there exist dedicated platforms making use of external automatic theorem
provers (ATP), or even some automatic tactics from interactive theorem
provers (ITP), to perform symbolic reasoning, and for example to
discharge some subgoals in the proofs.
Tools like Rodin~\cite{deharbe2013,deharbe2014,abrial2007} have already integrated several provers, like Z3, as modules for proving the proof obligations generated from the model inputed by user.
The prover we used, also is Z3, is developed by Microsoft Research based on the satisfiability modulo
theories (SMT), mainly applied in extended static checking, test case
generation, and predicate abstraction.
In a similar way, there are several ATPs/ITPs we could consider to use for
the result pruning and bisimulation checking in our algorithm, as an
alternative to Z3, such as CVC4~\cite{barrett:CAV2011},
Coq~\cite{armand:CPP2011},
%Coq~\cite{armand:CPP2011,besson:CPP2011},
Isabelle~\cite{blanchette:FroCoS2011} or others. 
}

@@

Several publications \cite{HMZ:PDP15,henrio:Forte2016} already have introduced
many examples of pNets, encoding
operators of various classical process algebras, or more complex
synchronization structures in distributed or parallel languages.
In this paper, we use a running example inspired from Lotos, to illustrate
the main features of pNets and the way we deal with them in the symbolic
semantics, and in the satisfiability checking.

\begin{figure}[h]
  \centerline{\includegraphics[width=8cm]{XFIG/LOTOSFormula.eps}}
  \caption{pNet encodings for the LOTOS expression}  \label{schema:lotos-pnet}
\end{figure}

@@

We use a brute-force method to generate all the possible open
transitions in the open automaton, using all possible combinations of
synchronization vectors. Naturally this builds some transitions where
the predicates express incompatible constraints. Even if having an
unsatisfiable (symbolic) transition in the open automaton would not be incorrect,
we want to check them for satisfiability, and reduce the number of
transitions and states in the automaton.
In Fig.~\ref{schema:unsat-ot}, we display an example of an
unsatisfiable open transition 
from the result of our example. It shows the case where
the controller C1 wants to move the control from $P$ to $(acc(x);Q)$,
conducting a $\delta$ transition. However, a synchronization
vector is chosen that does not match with this behavior, and we can
easily find the contradiction in the 
generated predicate term ``$\OTvar{:ra:11:1}=l \!\land\! \delta=\OTvar{:ra:11:1}$''.

@@

\begin{figure}[t]
  \centerline{\includegraphics[width=11cm]{XFIG/SMTLIB2}}
  \caption{The input of the Z3 solver in SMT-LIB language and the output result}  \label{schema:smt-lib}
\end{figure}

@@

\QIN{The $ot_1$ shows that only the actions of the process $P$ could be transmitted out before the value $x$ being captured which is represented by the $ot_2$. The other 3 open transitions shows the behaviors of the process $Q$ while the $ot_3$ and $ot_4$ won't be performed as the controller has already set the value $s_0$ to 1 in the previous action.
}

\QIN{
To sum up, from the pNet encoding LOTOS formula inputed by the user we generate the open transitions and prune the unsatisfiability automatically. The final result open transitions represent all the possible movements of the LOTOS example. They can exactly construct the open automaton for the given example.
}

>>>>

%\usepackage{amsfonts}
%\usepackage{fancyhdr}
%\usepackage{cite}
%\usepackage{ifthen}
%\usepackage{amssymb}
%\usepackage{fancyhdr}
%\usepackage{pifont}
\usepackage{stmaryrd}
\usepackage{mathtools,mathpartir}
\usepackage{proof}
%\usepackage{setspace}
%\usepackage{indentfirst}
\usepackage{amsmath,amssymb,amscd,mathrsfs}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
\usepackage{epsfig,color,subfigure,enumitem}
\newcommand{\TODO}[1]{\textcolor{red}{\textbf{[TODO:#1]}}}
\newcommand{\NOTE}[1]{\textcolor{blue}{\textbf{[NOTE:#1]}}}
\newcommand{\ERIC}[1]{\textcolor{blue}{#1}}
\newcommand{\OTvar}{\texttt}
\newcommand{\OTland}{\;\land\ }
\definecolor{airforceblue}{rgb}{0.26, 0.44, 0.56}
\newcommand{\QIN}[1]{\textcolor{airforceblue}{#1}}
\newcommand{\coloncolon}{{:\hspace{-.2ex}:}}
\makeatletter
\newcommand{\raisemath}[1]{\mathpalette{\raisem@th{#1}}}
\newcommand{\raisem@th}[3]{\raisebox{#1}{$#2#3$}}
\makeatother

@@

\title{Using SMT engine to generate Symbolic Automata\thanks{This work was partially 
funded by the Associated Team FM4CPS
  between INRIA and ECNU, Shanghai}}
\author{ Xudong Qin\inst{2,3}  \ \ \  Eric Madelaine\inst{1,2}
  \ \ \  Min Zhang\inst{3}} 
\institute{Univ. of Nice Sophia Antipolis, CNRS, UMR 7271, 06900 Sophia Antipolis, France
	\and INRIA Sophia Antipolis M\'edit\'erann\'ee, BP 93, 06902 Sophia Antipolis, France
\and Shanghai Key Laboratory of Trustworthy Computing, ECNU, China}
\date{}                                           % Activate to display a given date or no date

@@

\begin{abstract}
  %% \TODO{not good for a conference, (to be rewriten by Eric)}
  %% We implement the symbolic semantics of open pNets using their
  %% so-called ``Open Automaton'' behavioral semantics. This involve
  %% building predicates expressing the synchronization conditions
  %% allowing some combination of events in the pNet system. These
  %% predicates are typically built using first order logic, plus some
  %% predicates specific of particular action algebras. To reduce the
  %% complexity of the generated open automata, we use the Z3 SMT engine
  %% to check satisfiability of the predicates, and prune the state
  %% space..

  Open pNets are used to model the behavior of open
  synchronous or asynchronous systems expressed in various calculi
  or languages. They are endowed with a symbolic operational semantics
  in terms of so-called ``Open Automata''. We implement an algorithm
  computing this semantics, and building predicates expressing the
  synchronization conditions allowing some combination of events in
  the pNet system. Checking such predicates requires symbolic
  reasoning over first order logics, and specific functions of each
  action algebra. To reduce the complexity of the generated open
  automata, we use the Z3 SMT engine to check satisfiability of the
  predicates, and minimize the number of significant symbolic
  transitions.

@@

Parameterized Networks of Synchronized Automata (pNets) were proposed
to give a behavioral specification formalism for distributed
systems, synchronous, asynchronous, or heterogeneous. It is used in
VerCors, a platform for designing and 
verifying distributed systems, as the intermediate language for various
high-level languages. The high-level languages in VerCors formalize
each component of the 
distributed system and gives out the composition of these
components.
pNets provides the core low-level semantic formalism for VerCors, and
is made of a hierarchical composition of (value-passing) automata,
called parameterized labelled transition systems (pLTS), where each
hierarchical level defines the possible synchronization of the lower levels.
Traditionally, pNets have been used to formalize fully
defined systems or softwares. But we want also to define and reason
about incompletely defined systems, like program skeletons, operators,
or open expressions of process calculi.
The open pNet is proposed to solve the
problem of the undefined components contained in the systems using
"holes" as the process parameter dealing with these components. The
hole acts as the placeholder for the uncertain component with a set of
possible behaviors the component might conduct, presenting the "open"
property.  
The pNet model was developed in a series of
papers~\cite{HMZ:PDP15,henrio:Forte2016} and some examples have been
introduced showing its ability to encode the operators from some
other algebras or  program skeletons. The operational semantic of the open pNet has
already been defined, but we only had a sketch about how to compute it and
had no implementation. This operational semantic is defined as an
Open Automaton in which Open Transitions contain logical predicates
expressing the relations between the behavior of the holes, and the
global behavior of the system.

Computing the open transitions from the composition of pNet is the
main part of the algorithm and needs to satisfy the constraints expressed as
the rules of the operational semantics.
Computing open automata by hand is complicated and error prone. And
the algorithm published in \cite{henrio:Forte2016} was not precise
enough for an implementation. Our first goal was then to refine the
algorithm, and implement it in a tool.
In the long goal, we want to be able to check the equivalence between
pNets. The equivalence between pNets is "FH-bisimulation" taking the
predicate of the open transitions into account each time matching such
open transitions. In our long-term goals, the final target is to check
the property of open systems.

@@

\paragraph{Contribution}
In the article we show how:
\begin{itemize}
\item We have refined the open automaton generation algorithm, to make
  it implementable, and we implemented a full working prototype, within the
  VerCors platform. In the process, we had to improve significantly the
  semantics rules, and add features in the algorithm to deal the full
  model, including management of variables and assignments.
%  \item We extended the meta-model of action expressions, and the
%    associated API, in the VerCors platform, to be able to manage
%    parameterisation over the action algebra operators.
  \item We have implemented the interaction between our algorithm and the Z3
    SMT solver, for checking satisfiability of the logical predicates
    generated by the algorithm.
    \item We validated the whole approach on a small use-case covering all
      important features of the model.
\end{itemize}

@@

@@

For other applications, e.g. for the analyses of 
programming languages, there exist dedicated platforms making use of
external automatic theorem 
provers (ATP), or even some automatic tactics from interactive theorem
provers (ITP), to perform symbolic reasoning, and for example to
discharge some subgoals in the proofs.
Tools like Rodin~\cite{deharbe2013,deharbe2014,abrial2007} have
already integrated several provers, like Z3, as modules for proving
the proof obligations generated from the model inputed by user. 
The prover we used, also is Z3, is developed by Microsoft Research
based on the satisfiability modulo 
theories (SMT), mainly applied in extended static checking, test case
generation, and predicate abstraction.
In a similar way, there are several ATPs/ITPs we could consider to use for
the result pruning and bisimulation checking in our algorithm, as an
alternative to Z3, such as CVC4~\cite{barrett:CAV2011},
Coq~\cite{armand:CPP2011},
%Coq~\cite{armand:CPP2011,besson:CPP2011},
Isabelle~\cite{blanchette:FroCoS2011} or others.

@@

Several publications \cite{HMZ:PDP15,henrio:Forte2016} already have introduced
many examples of pNets, encoding
operators of various classical process algebras, or more complex
synchronization structures in distributed or parallel languages.
In this paper, we use an example inspired from Lotos, to illustrate
the main features of pNets and the way we deal with them in the symbolic
semantics, and in the satisfiability checking.

\begin{figure}[h]
  \centerline{\includegraphics[width=10cm]{XFIG/LOTOSFormula.eps}}
  \caption{pNet encodings for the LOTOS expression}  \label{schema:lotos-pnet}
\end{figure}

@@

We use a brute-force method to generate all the possible open
transitions in the open automaton, using all possible combinations of
synchronization vectors. Naturally this builds some transitions where
the predicates express incompatible constraints. Even if having an
unsatisfiable (symbolic) transition in the open automaton would not be incorrect,
we want to check them for satisfiability, and reduce the number of
transitions and states in the automaton.
In Fig.~\ref{schema:unsat-ot}, we display an example of an
unsatisfiable open transition 
from the result of our example. It shows the case where
the controller C1 wants to move the control from $P$ to $(acc(x);Q)$,
conducting a $\delta$ transition. However, a synchronization
vector is chosen that does not match with this action, and we can
easily find the contradiction in the 
generated predicate term ``$\OTvar{:ra:11:1}=l \!\OTland\! \delta=\OTvar{:ra:11:1}$''.

@@

\begin{figure}[t]
  %  \centerline{\includegraphics[width=12cm]{XFIG/SMTLIB2}}
    \centerline{\includegraphics[width=\linewidth]{XFIG/SMTLIB2}}
  \caption{The input of the Z3 solver in SMT-LIB language and the output result}  \label{schema:smt-lib}
\end{figure}

@@

The $ot_1$ transition means that the actions of the process $P$ could
be transmitted out, until $P$ performs a $\delta(x)$ and the value of
$x$ is transmitted, through transition
$ot_2$. The other 3 open transitions shows the behaviors of the
process $Q$ while the $ot_3$ and $ot_4$ won't be performed as the
controller has already set the value $s_0$ to 1 in the previous
action.

To sum up, from the pNet encoding LOTOS expression, inputed by the user, we
generate the open transitions and prune the unsatisfiability
automatically. The final result open transitions represent all the
possible movements of the LOTOS example. They can exactly construct
the open automaton for the given example.
